---
title: "Base editing design with crisprDesign"
author: 
- name: Jean-Philippe Fortin
  email: fortin946@gmail.com
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc_float: true
    #theme: paper
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Base editing design with crisprDesign}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
---



# Introduction




```{r, message=FALSE, warning=FALSE,results='hide' }
library(crisprDesign)
```









# Defining the base editor object


We illustrate the CRISPR base editing (CRISPRbe) functionalities 
of `crisprDesign` by designing and characterizing gRNAs targeting the gene
IQSEC3 using the cytidine base editor BE4max [@koblan2018improving]. 

We first load the BE4max `BaseEditor` object available in `crisprBase`:

```{r}
data(BE4max, package="crisprBase")
BE4max
```

The editing probabilities of the base editor BE4max are stored in a matrix 
where rows correspond to the different nucleotide substitutions, and columns
correspond to the genomic coordinate relative to the PAM site. 
The `editingWeights` function from `crisprBase` allows us to retrieve 
those probabilities. One can see that C to T editing is optimal 
around 15 nucleotides upstream of the PAM site for the BE4max base editor:

```{r}
crisprBase::editingWeights(BE4max)["C2T",]
```

To learn how to build a \code{BaseEditor} object for your own base editor
enzyme, see the \code{crisprBase} package vignette.


# Designing spacer sequences

Next, we load the
`grListExample` object in `crisprDesign` that contains gene 
coordinates in hg38 for exons of all human IQSEC3 isoforms. The object was
obtained by converting an Ensembl `TxDb` object into a `GRangesList`
object using the `TxDb2GRangesList` convenience function in `crisprDesign`.


```{r}
data(grListExample, package="crisprDesign")
```


The `queryTxObject` function allows us to query such objects for a specific
gene and feature. Here, we obtain a `GRanges` object containing the first exon
of the IQSEC3 gene:


```{r}
gr <- queryTxObject(txObject=grListExample,
                    featureType="cds",
                    queryColumn="gene_symbol",
                    queryValue="IQSEC3")
```

and retain the first exon only:

```{r}
gr <- gr[1]
```

`findSpacers` is the main function to obtain a list of all
possible spacer sequences targeting protospacers located in the target
DNA sequence(s). If a `GRanges` object is provided as input, a `BSgenome`
object (object containing sequences of a reference genome) will need to be
provided as well:

```{r, warning=FALSE, message=FALSE}
library(BSgenome.Hsapiens.UCSC.hg38)
bsgenome <- BSgenome.Hsapiens.UCSC.hg38
guideSet <- findSpacers(gr,
                        bsgenome=bsgenome,
                        crisprNuclease=BE4max,
                        strict_overlap=FALSE)
guideSet
```


The argument \code{strict_overlap} set to FALSE enables spacer sequences to be
not-strictly overlapping with the CDS region; this is useful for base editing
design as the editing window can extend beyond the protospacer sequence region.
The function returns a `GuideSet` object that stores genomic coordinates for all spacer
sequences found in the regions provided by `gr`. The `GuideSet` object is an
extension of a `GenomicRanges` object that stores additional information about
gRNAs. For the sake of time, we will retain only two gRNAs:



```{r}
guideSet <- guideSet[c(50,51)]
print(guideSet)
```


# Allele prediction

The function `addEditedAlleles` finds, characterizes, and scores predicted
edited alleles for each gRNA, for a chosen transcript. It requires a 
transcript-specific annotation that can be obtained using the 
function `getTxInfoDataFrame`. Here, we will perform the
analysis using the main isoform of IQSEC3 (transcript id ENST00000538872).

We first get the transcript table for ENST00000538872, 

```{r}
txid <- "ENST00000538872"
txTable <- getTxInfoDataFrame(tx_id=txid,
                              txObject=grListExample,
                              bsgenome=bsgenome,
                              extend=30)
head(txTable)
```

The argument `extend` specifies the number of nucleotides upstream and downstream
of the exons to include. This is useful to characterize gRNAs overlapping
splice junctions. The `region` column indicates where the location of the 
nucleotide: 3UTR, 5UTR, CDS, Intron, and Upstream and Downstream of the CDS.


We are ready to add predicted alleles to the `GuideSet` object:

```{r}
editingWindow <- c(-20,-8)
guideSet <- addEditedAlleles(guideSet,
                             baseEditor=BE4max,
                             txTable=txTable,
                             editingWindow=editingWindow,
                             minEditingWeight = 0,
                             minMutationScore = 0.3)
```

- The `editingWindow` argument specifies the window of editing that
we are interested in. When not provided, it uses the default window
provided in the `BaseEditor` object. Note that providing large windows 
can exponentially increase computing time as the number of possible 
alleles grows exponentially.

- The `minEditingWeight` specifies the minimum editing weight required for an
allele to be listed as a predicted allele. Default is 0. A higher threshold can
be used to omit alleles with low probabilities.

- The `mutationScore` specifies a minimum predicted probability for labeling an 
allele with a predicted variant. Alleles with scores lower than this threshold
will be labeled as "not_editing". Default of 0.3.

For each gRNA, a predicted alleles annotation is stored and can be retrieved
using the `editedAlleles` accessor function. As an example, let's retrieve
the predicted alleles for the first gRNA:


```{r}
alleles <- editedAlleles(guideSet[1])
print(alleles)
```


The resulting `DataFrame` is ordered so that the top predicted alleles 
(based on the `score` column) are shown first. The `score` 
represents the likelihood of the edited allele to occur relative
to all possible edited alleles, and is calculated using the editing
weights stored in the `BE4max` object. The `seq` column represents 
the edited nucleotide sequences. They are always reported from the 5' to 3' 
direction on the strand corresponding to the gRNA strand. 

- The `n_mismatches` column indicates the number of amino acid that differs
between the edited allele and the wildtype allele.
- The `n_nonsense` and `n_missense` columns indicate the number of mismatches
that are nonsense and missense mutation, respectively. Those two columns sum to 
the `n_mismatches` column. 
- The `variant` column indicates the functional consequence of the allele.
There are 7 possible choices:
    - `silent`: single or multiple silent mutations
    - `missense`: single missense mutation
    - `nonsense`: single nonsense mutation
    - `missense_multi`: multiple missense mutations
    - `nonsense_multi`: multiple nonsense mutations
    - `splice_junction`: mutation in a splice junction
    - `not_targeting`: no mutations found in CDS or splice junctions
In case an edited allele leads to multiple editing events that have different
variant label, the most detrimental mutation
(splice junction over nonsense, nonsense over missense,
missense over silent) is reported. 
- The `positions` column lists the amino acid numbers where mutations occur. 
For alleles that are labeled as `splice_junction`, it lists the closest amino 
acid. 
- The `aa` column reports the result edited amino acid sequence. 


The alleles object also contains useful metadata information that can be
accessed using the `metadata` accessor function:

```{r}
metadata(alleles)
```

The `wildtypeAllele` reports the unedited nucleotide sequence of the
region specified by the editing window (with respect to the gRNA PAM site).
It is always reported from the 5' to 3' direction on the strand corresponding 
to the gRNA strand. The `start` and `end` specify the corresponding 
coordinates on the transcript. 


# gRNA-level aggregate variant scores

For both analysis and visualization purposes, it is often useful to label gRNAs 
with a score and label that represents the most likely functional consequence
of that gRNA, for a given base editor. The `addEditedAlleles` function described
above also implements a gRNA-level aggregate scoring scheme that adds 
several gRNA-level aggregate scores to the `GuideSet` object:

```{r}
head(guideSet)
```

- `score_missense_single`: sum of probability scores across all alleles with a single missense mutation
- `score_nonsense_single`: sum of probability scores across all alleles with a single nonsense mutation
- `score_missense_multi`: sum of probability scores across all alleles with multiple missense mutations
- `score_nonsense_multi`: sum of probability scores across all alleles with multiple nonsense mutations
- `score_silent`: sum of probability scores across all alleles with only silent mutations
- `score_splice_junction`: sum of probability scores across all alleles with a splice junction mutation
- `score_missense`: `score_missense_single` and `score_missense_multiple` added together
- `score_nonsense`: `score_nonsense_single` and `score_nonsense_multiple` added together
- `maxVariantScore`: maximum score across all score columns
- `maxVariant`: variant label for the maximum score
- `aapos`: amino acid position for the top predicted allele for the variant category that has the maximum score

For both gRNAs, the highest scores are missense, and therefore the `maxVariant` is set to missense. 

# Allele scoring algorithm in details





# Session Info

```{r}
sessionInfo()
```

# References

